KIN: A method to infer relatedness from low-coverage ancient DNA

### 2022-09-08 paper_overleaf/main.tex

\begin{figure}[h!]
    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic_sib_rev.png}
    \centering
    \caption{IBD sharing between siblings without and with runs of homozygosity (ROH). (A) Schematic of chromosomes for parent 1 (Par1)and parent 2 (Par2). On the left, the parents have unrelated chromosomes depicted with different colors, and on the right parent 1 has a homozygous pair of chromosomes. (B) In top row, recombined chromosomes inherited from parents by a pair of siblings are shown. The bottom row shows the expected proportion of differences ($p$) along the chromosome. In both cases, $p$ can take values of $p_0$, $p_1$ or $p_2$, which are the expected proportion of differences in IBD states 0,1 and 2, respectively.  }
    \label{fig0:schematic}
\end{figure}

@@

\subsection{Input of KIN-HMM}\label{hmm_input}
The inputs of our algorithm are i) the number of overlapping sites for the $w$-th window $N_w$ for which both samples have at least one read available, ii) the expected number of pairwise differences $D_w$ at these sites, and iii) the probability of ROH in windows, by default obtained from ROH-HMM described in section \ref{roh}.

$D_w$ is the number of sites that are different between a pair of individuals. For high-coverage data, this can be estimated accurately by comparing genotypes, but for low-coverage samples, $D_w$ needs to be estimated. The simplest approach is to randomly-sample a read from each position \cite{haak_massive_2015,prufer_complete_2014,green_draft_2010}. However, such an approach may result in loss of data, and hence we calculate $D_w$ by implicitly summing over all possible samplings:
\begin{align}\label{eq:x}
D_w &= \sum_{s=1}^{N_w} \nu_i(s) (1-\nu_j(s)) + (1-\nu_i(s)) \nu_j(s)
\end{align}
Here, $\nu_i(s)$ and $\nu_j(s)$ are the proportions of reads carrying the derived allele at SNP index $s$ for individuals i and j respectively.
Throughout, we will use bold-face notation to refer to the vector (or matrix) collecting  all the terms, e.g. $\BD = (D_1, D_2, \dots D_L)$.

@@

There are three additional model parameters, $\pi, \mathbf{A}$ and $\delta$. The initial probabilities $\pi$ give the probabilities of being in each state $Z_0$ (at the beginning of each chromosome), which we set to uniform for simplicity. The transition matrix $\mathbf{A}$ gives the probability of moving from state $i$ to state $j$, given by $a_{ij}$, and is fixed and estimated from simulations for each relatedness case (section \ref{simulat}). The overdispersion-parameter $\delta$ takes into account that SNPs in each window vary in their allele frequency (see next setion).  For compactness of notation, we group the fixed parameters: $\theta = (\BN, \mathbf{A}, \pi)$.

@@

The presence of long tracts of homozygosity resulting from recent inbreeding adds an additional complication, as the number of shared chromosomes may be overestimated \cite{weir_genetic_2006}. For example, when considering two bones from the same individual, we would expect the entire genome to have a pairwise difference of $p_2$, because two out of the four compared chromosomes are identical copies. However, in inbred regions all four chromosomes will be identical, and so the expected pairwise differences are zero ($p_3$ in equation \ref{eq:2}). Note however, that $p_0$ does not depend on the presence of ROH, since all comparisons are between unrelated chromosomes even if both individuals are homozygous at a particular locus.

Taken together, we can summarize $p$ in the following matrix, where rows give the state of $Z_w$, and columns of $H_w$:
\begin{equation}\label{eq:2}
    p(Z_w, H_w) = \left[\begin{array}
{rrr}
p_0 & p_0 & p_0 \\
p_1 & p_2 & p_3 \\
p_2 & p_3 & p_3
\end{array}\right]
\end{equation}

As explained above, we would expect $p_3$ to be zero. However, as we do all our calculations in large windows, the start/end positions of windows may not coincide with that of ROH tracts, and we found that we obtain better results by setting $p_3 = \frac{p_2}{2}$.

@@

Using equation \ref{eq:2}, we simplify this by grouping all the terms that would result in the same $p_i$ and $\delta_i$, i.e.

\begin{align*}\label{eq:5}
g_{w0} &= \gamma_{w0}\\
g_{w1} &= \gamma_{w1} h_{w0}\\
g_{w2} &= \gamma_{w2} h_{w0} + \gamma_{w1} h_{w1}\\
g_{w3} &= \gamma_{w2} h_{w1} + \gamma_{w2} h_{w2} + \gamma_{w1} h_{w2}
\end{align*}

So we can rewrite equation \ref{eq:4} as:
\begin{align}\label{eq:6}
\mathcal{C} &= \sum_{i=0}^3 \sum_{w=1}^L \log P(D_{w}|N_w, \delta_i, p_i)g_{wi}
\end{align}

@@

\subsection{Model comparison}\label{model_comp}
To infer the most likely relatedness case, we run our model on all relatedness cases mentioned in section \ref{method_overview}, and compare the resulting likelihoods. We output the relatedness corresponding to the maximum likelihood model. Because the models we test are not nested, standard likelihood-ratio theory cannot be used to generate a confidence value. Instead we use the log likelihood ratio between the two best models as a statistic to assess the confidence in our classifications, and use simulations to obtain critical values (Fig. S8).

@@

\paragraph{Example case of siblings:}
In Fig.\ref{fig1:ibd} we show the inferred IBD fragments from different KIN-HMMs when they are applied to simulated data from a pair of siblings. In this case, the siblings-model predictions match true IBD states the most, as reflected by the highest log-likelihood.

\begin{figure}[h!]
    \includegraphics[width=16cm]{plots/inkscape_finalImg/IBDplot_title.png}
    \centering
    \caption{Comparison of pairwise difference data and inferred IBD fragments. The top panel shows the proportion of differences in each window along the genome for a pair of simulated siblings. Dashed lines represent $p_0$, $p_1$ and $p_2$ estimates. The second panel shows the true IBD state for each window. The remaining panels show the IBD states predicted by particular relatedness models. The log likelihood value for each model is shown on upper left corner of the panel. Light and shaded background represent distinct chromosomes.}
    \label{fig1:ibd}
\end{figure}

@@

Since the source of ROH may not be known, we estimate both transitions and emissions. We calculate the emissions using a beta-binomial likelihood, and fix the mean of the distributions corresponding to $Y_w$ = 3 and $Y_w$ = 2 at expected proportion of differences in a homozygous tract ($p_3$) and expected proportion of differences in a non-homozygous tract ($p_2$) respectively. The expectation step outputs the posterior probability $\Gamma$ of being in state $Y_w=3$ or the state $Y_w=2$ in each window. The maximization step for emissions is analogous to that in the KIN model (eq. \ref{eq:4}), and the optimization step here is done with the following cost function:

@@

\paragraph{IBD prediction:}
We investigated the accuracy of IBD state prediction along the genome by counting the number of genomic windows where we correctly predict IBD state for different relatedness and coverages (Fig. \ref{fig4:IBDstate_accuracy}). The accuracy for coverages of 0.1x or higher is consistent and varies between 1.0 and 0.836, depending on relatedness. However, the accuracy decreases at lower coverages for most relatedness cases and ranges from 1.0 to 0.563 at 0.03x. The exceptions are for identical individuals and parent-child pairs, where the accuracy is nearly perfect for all investigated coverages, even at 0.03x coverage. We find that adding contamination, ascertainment bias in our simulations has little effect on IBD prediction. Adding ROH to the simulations reduces the IBD prediction in two cases: the accuracy for second degree is decreased by $1.5$ to $5\%$ and for siblings by $5$ to $17\%$ (Fig. S9). We see this adverse effect of ROH in case of siblings, perhaps because only in case of siblings we do not introduce long ROH directly, but long and short ROH may be present through parents (see section \ref{method}). Unidentified short ROH may cause a difficulty in differentiating between different combinations of IBD ($Z_w$) and ROH states ($H_w$). However, this does not affect the power to identify siblings in presence of ROH even at $0.05x$ (see Fig. \ref{fig3:Comparison_READ_KIN})

\begin{figure}[h!]
    \includegraphics[width=16cm]{plots/plotimg/plot_IBDaccuracy.png}
    \centering
    \caption{Evaluation of IBD estimation at different coverages. y-axis shows accuracy calculated over 60 simulations with each relatedness case and six different coverages shown on x-axis. The error bars are drawn at 1 standard deviation from the mean. Here, accuracy corresponding to relatedness cases for Parent-Child and Identical individuals is always 1, and overlaps with each other. Accuracy is defined as the sum of true positives and true negatives divided by the total number of windows. We consider a prediction is true if it matches the IBD state at the central position of the window.}
    \label{fig4:IBDstate_accuracy}
\end{figure}

@@

We compared the IBD estimates obtained using KIN to those from lcMLkin for all pairs for whom READ results (with s.d. $>$1) and KIN results ($\Delta LL>$1) match (Fig. S10). We find that lcMLkin creates four different clusters for different relatedness cases (identical, parent-child, $2^{nd}$ degree and unrelated). However, the location of these clusters strongly deviate from the expected values, likely due to the low coverage and ROH. Remains from the same individual, for example, are expected to be at $k_0 = 0$ and $r = 1$, but are at $k_0 \approx 0.3$, $r \approx 0.6$.

@@

Only for 28 pairs, KIN has confident classifications that differ from those obtained using READ (table S2, Fig. S11). For 20 of these comparisons, KIN predicts a third-degree relationship, which READ classifies as unrelated.

lcMLkin predicts $3^{rd}$ - $5^{th}$ in 14 of these cases, unrelated in 1 case (see Fig. S11), second degree in 1 case (see Fig. S11), and does not have enough data for classification in 4 cases. For 10 additional pairs, lcMLkin predicts $3^{rd}$ - $5^{th}$ degree, but KIN infers them to be unrelated.

@@

For detecting ROH, there is only one method available that works with low coverage ancient DNA \cite{ringbauer_parental_2021}. This software can infer ROH at coverages $\geq0.3x$, but requires a reference panel. In the absence of a method to detect long ROH at low coverages without a reference panel, we developed a HMM based method, and tested it on simulations to show that it can be used reliably at coverage $\geq 0.1x$ to detect ROH regions with lengths on the order of 10cM. 
In our simulations, we find that adding long ROH ($~17\%$) to simulations slightly improves the power of both KIN and READ, particularly at lower coverages. This is likely because ROH actually reduces the variance in differences depending on the relatedness, and makes it easier to correctly classify relatedness. For READ, the presence of ROH causes a bias towards inferring closer relatedness cases, but the model we use in KIN reduces this bias (see Fig. \ref{fig3:Comparison_READ_KIN}).

>>>>

\begin{figure}[h!]
    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic_sib_rev.png}
    \centering
    \caption{IBD sharing between siblings without and with runs of homozygosity (ROH). (A) Schematic of chromosomes for two parents (Par1, Par2). (B) Schematics of recombinant chromosomes of two children (Sib1, Sib2). (C)  Expected differences between Sib1 and Sib2 ($p$) along the chromosome. In both cases, $p$ can take values of $p_0$, $p_1$ or $p_2$, which are the expected proportion of differences in IBD states 0,1 and 2, respectively.  (D-F): Same as A-C, except parent 1 is assumed to be homozygous.  }
    \label{fig0:schematic}
\end{figure}

@@

\subsection{Input of KIN-HMM}\label{hmm_input}
The inputs of our algorithm are i) the number of overlapping sites for the $w$-th window $N_w$ for which both samples have at least one read available, ii) the number of pairwise differences $D_w$ at these sites, and iii) the probability of ROH in windows, by default obtained from ROH-HMM described in section \ref{roh}.

For high-coverage data, $D_w$ can be directly obtained by comparing genotypes, but for low-coverage samples, $D_w$ needs to be estimated from the sequencing data. The simplest approach is to randomly-sample a read from each position \cite{haak_massive_2015,prufer_complete_2014,green_draft_2010}. However, such an approach may result in loss of data, and hence we estimate $D_w$ by implicitly summing over all possible samplings:
\begin{align}\label{eq:x}
D_w &= \sum_{s=1}^{N_w} \nu_i(s) (1-\nu_j(s)) + (1-\nu_i(s)) \nu_j(s)
\end{align}
Here, $\nu_i(s)$ and $\nu_j(s)$ are the proportions of reads carrying the derived allele at SNP index $s$ for individuals i and j respectively.
Throughout, we will use bold-face notation to refer to the vector (or matrix) collecting  all the terms, e.g. $\BD = (D_1, D_2, \dots D_L)$.

@@

There are three additional model parameters, $\pi, \mathbf{A}$ and $\delta$. The transition matrix $\mathbf{A}$ gives the probability of moving from state $i$ to state $j$, given by $a_{ij}$, and is fixed and estimated from simulations for each relatedness case (section \ref{simulat}). The initial probabilities $\pi$ give the probabilities of being in each state $Z_0$ (at the beginning of each chromosome), which we set to the stationary distribution of the transition matrix for simplicity. The overdispersion-parameter $\delta$ takes into account that SNPs in each window vary in their allele frequency (see next setion).  For compactness of notation, we group the fixed parameters: $\theta = (\BN, \mathbf{A}, \pi)$.

@@

The presence of long tracts of homozygosity resulting from recent inbreeding adds an additional complication, as the number of shared chromosomes may be overestimated \cite{weir_genetic_2006}. For example, when considering two bones from the same individual, we would expect the entire genome to have a pairwise difference of $p_2$, because two out of the four compared chromosomes are identical copies. However, in inbred regions all four chromosomes will be identical, and so the expected pairwise differences are zero ($p_4$ in equation \ref{eq:2}). Note however, that $p_0$ does not depend on the presence of ROH, since all comparisons are between unrelated chromosomes even if both individuals are homozygous at a particular locus.

Taken together, we can summarize $p$ in the following matrix, where rows give the state of $Z_w$, and columns of $H_w$:
\begin{equation}\label{eq:2}
    p(Z_w, H_w) = \left[\begin{array}
{rrr}
p_0 & p_0 & p_0 \\
p_1 & p_2 & p_4 \\
p_2 & p_4 & p_4
\end{array}\right]
\end{equation}

As explained above, we would expect $p_4$ to be zero. However, as we do all our calculations in large windows, the start/end positions of windows may not coincide with that of ROH tracts, and we found that we obtain better results by setting $p_4 = \frac{p_2}{2}$, to take into account that many windows will only partially have four comparisons between identical chromosomes.

@@

Using equation \ref{eq:2}, we simplify this by grouping all the terms that have the same number of pairwise comparisons between identical chromosomes, which would result in the same $p_i$ and $\delta_i$, i.e.

\begin{align*}\label{eq:5}
g_{w0} &= \gamma_{w0}\\
g_{w1} &= \gamma_{w1} h_{w0}\\
g_{w2} &= \gamma_{w2} h_{w0} + \gamma_{w1} h_{w1}\\
g_{w3} &= 0 \\
g_{w4} &= \gamma_{w2} h_{w1} + \gamma_{w2} h_{w2} + \gamma_{w1} h_{w2},
\end{align*}

where $g_{w3}$ is always 0 because there is no case that leads to only three comparisons between identical chromosomes.

So we can rewrite equation \ref{eq:4} as:
\begin{align}\label{eq:6}
\mathcal{C} &= \sum_{i=0}^4 \sum_{w=1}^L \log P(D_{w}|N_w, \delta_i, p_i)g_{wi}
\end{align}

@@

\subsection{Model comparison}\label{model_comp}
To infer the most likely relatedness case, we run our model on all relatedness cases mentioned in section \ref{method_overview}, and compare the resulting likelihoods. We output the relatedness corresponding to the maximum likelihood model. Since we compare models where the  parameters are not subset of each other, standard likelihood-ratio theory for nested models cannot be used to obtain confidence intervals. Instead we use the log likelihood ratio between the two best models as a statistic to assess the confidence in our classifications, and use simulations to obtain critical values (Fig. S8).

@@

\paragraph{Example case of siblings:}
In Fig.\ref{fig1:ibd} we show the inferred IBD fragments from different KIN-HMMs when they are applied to simulated data from a pair of siblings. The models for identical, parent-child or unrelated relationships allow for just one IBD state, resulting in a flat line and low likelihood for this data. The other three cases all allow for different IBD-states, but the siblings-model predictions match true IBD states the most, as reflected by the highest log-likelihood and the close correspondence of the inferred and true IBD states.

\begin{figure}[h!]
    \includegraphics[width=16cm]{plots/inkscape_finalImg/IBDplot_title.png}
    \centering
    \caption{Comparison of pairwise difference data and inferred IBD fragments. The top panel shows the proportion of differences in each window along the genome for a pair of simulated siblings. Dashed lines represent $p_0$, $p_1$ and $p_2$ estimates. The second panel shows the true IBD state for each window. The remaining panels show the IBD states predicted by particular relatedness models. The log-likelihood value for each model is shown on upper left corner of the panel. Light and shaded backgrounds represent distinct chromosomes.}
    \label{fig1:ibd}
\end{figure}

@@

Since the source of ROH may not be known, we estimate both transitions and emissions. We calculate the emissions using a beta-binomial likelihood, and fix the mean of the distributions corresponding to $Y_w$ = 3 and $Y_w$ = 2 at expected proportion of differences in a homozygous tract ($p_4$) and expected proportion of differences in a non-homozygous tract ($p_2$) respectively. The expectation step outputs the posterior probability $\Gamma$ of being in state $Y_w=3$ or the state $Y_w=2$ in each window. The maximization step for emissions is analogous to that in the KIN model (eq. \ref{eq:4}), and the optimization step here is done with the following cost function:

@@

\begin{table}[h!]
\caption{\label{tab:Table 3}Model performance for ROH prediction. Here we test ROH-HMM in four different cases of simulations: Control (without ROH, ascertainment bias or contamination), R (with ROH), RA (with ROH and ascertainment bias), RC (with ROH and contamination)}
\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
    \hline
    Coverage & R(Sensitivity) & R(Specificity) & RA(Sensitivity) & RA(Specificity) & RC(Sensitivity) & RC(Specificity)\\
    \hline
    4x & 0.007 & 0.006 & 0.087 & 0.038 & 0.089 & 0.039 & 0.101 & 0.042\\
    \hline
    0.5x & 0.006 & 0.005 & 0.085 & 0.036 & 0.087 & 0.036 & 0.098 & 0.039\\
    \hline
    0.2x & 0.004 & 0.003 & 0.073 & 0.03 & 0.073 & 0.028 & 0.084 & 0.031\\
    \hline
    0.1x & 0.003 & 0.002 & 0.054 & 0.023 & 0.054 & 0.023 & 0.065 & 0.025\\
    \hline
\end{tabular}
\label{table2}
\end{table}

\paragraph{IBD prediction:}
We investigated the accuracy of IBD state prediction along the genome by counting the number of genomic windows where we correctly predict IBD state for different relatedness and coverages (Fig. \ref{fig4:IBDstate_accuracy}). The accuracy for coverages of 0.1x or higher is consistent and varies between 1.0 and 0.78, depending on relatedness. However, the accuracy decreases at lower coverages for most relatedness cases and ranges from 1.0 to 0.48 at 0.03x. The exceptions are for identical individuals and parent-child pairs, where the accuracy is nearly perfect for all investigated coverages, even at 0.03x coverage. We find that adding contamination, ascertainment bias in our simulations has little effect on IBD prediction. Adding ROH to the simulations reduces the IBD prediction in two cases: the accuracy for second degree is decreased by $1.5$ to $5\%$ and for siblings by $5$ to $17\%$ (Fig. S9). We see this adverse effect of ROH in case of siblings, perhaps because only in case of siblings we do not introduce long ROH directly, but long and short ROH may be present through parents (see section \ref{method}). Unidentified short ROH may cause a difficulty in differentiating between different combinations of IBD ($Z_w$) and ROH states ($H_w$). However, this does not affect the power to identify siblings in presence of ROH even at $0.05x$ (see Fig. \ref{fig3:Comparison_READ_KIN})

\begin{figure}[h!]
    \includegraphics[width=16cm]{plots/plotimg/plot_IBDaccuracy.png}
    \centering
    \caption{Evaluation of IBD estimation at different coverages. y-axis shows accuracy calculated over 60 simulations with each relatedness case and six different coverages shown on x-axis. The error bars are drawn at 1 standard deviation from the mean. Here, accuracy corresponding to relatedness cases for Parent-Child and Identical individuals is always 1, and overlaps with each other. Accuracy is defined as the proportion of correctly predicted IBD states, when compared to the central position of the window.}
    \label{fig4:IBDstate_accuracy}
\end{figure}

@@

We compared the IBD estimates obtained using KIN to those from lcMLkin for all pairs for whom READ results (with s.d. $>$1) and KIN results ($\Delta LL>$1) match (Fig. S10). We find that lcMLkin creates four different clusters based on the coefficient of relatedness ($r$) and the proportion of the genome in the unrelated state ($k_0)$ for different relatedness cases (identical, parent-child, $2^{nd}$ degree and unrelated). However, the location of these clusters strongly deviate from the expected values, likely due to the low coverage and ROH. Remains from the same individual, for example, are expected to be at $k_0 = 0$ and $r = 1$, but are at $k_0 \approx 0.3$, $r \approx 0.6$ for lcMLkin (Figure S10).

@@

Only for 28 pairs, KIN has confident classifications that differ from those obtained using READ (table S2, Fig. S11). Twenty of these comparisons, KIN predicts to be third-degree relationships, and  READ concordantly classifies them as unrelated (as it only infers first and second degree relationships). lcMLkin predicts $3^{rd}$ - $5^{th}$ in 14 of these cases, unrelated in 1 case (see Fig. S11), second degree in 1 case (see Fig. S11), and does not have enough data for classification in 4 cases. For 10 additional pairs, lcMLkin predicts $3^{rd}$ - $5^{th}$ degree, but KIN infers them to be unrelated. There are a total of 83 pairs where READ obtains a confident call ($s.d.>1$), but KIN does not ($\Delta LL<1$). In 80 of these cases, READ classifies them as unrelated, KIN classifies $3^{rd}$ degree, while lcMLkin predicts $3^{rd}$ - $5^{th}$ degree. In rest 3 cases READ classifies unrelated, while KIN and lcMLkin agree on $2^{nd}$ degree.

@@

For detecting ROH, there is only one method available that works with low coverage ancient DNA \cite{ringbauer_parental_2021}. This software can infer ROH at coverages $\geq0.3x$, but requires a large reference panel. Instead, our method detects long ROH based on just the expected heterozygosity $p_0$, which can be estimated from a small number of unrelated individuals.  On simulations, we find that our method reliably inferrs ROH regions  with lengths on the order of 10cM from samples sequenced to coverage $\geq 0.1x$ . 
In our simulations, we find that adding long ROH ($~17\%$) to simulations slightly improves the power of both KIN and READ, particularly at lower coverages. This is likely because ROH actually reduces the variance in differences depending on the relatedness, and makes it easier to correctly classify relatedness. For READ, the presence of ROH causes a bias towards inferring closer relatedness cases, but the model we use in KIN reduces this bias (see Fig. \ref{fig3:Comparison_READ_KIN}).
